Phoneme surprisal

### 2019-07-01 surprisal/phoneme_writeup.Rmd
tidying phoneme writeup

```{r setup, include=FALSE}
library(tidytext)
library(tidyverse)
library(childesr)
library(glue)
library(feather)
library(here)
library(entropy)
library(tidyboot)
library(janitor)
library(arrangements)

knitr::opts_chunk$set(echo = TRUE)
```

We're going to discuss the amount of information across letters in English, then information in English phones (sounds), then the hold-out entropy in English and finally the information in letters in French and German. We are interested in French because when many French words are pronounced, the final letters are not pronounced in any way (e.g. "ils sauraient"="they would know", the final "ent" in "sauraient" is not pronounced at all). We are interested in German because German has a much richer case system than English or French, and case markings in German are suffixes (at the ends of words). So in German, the ends of words may be much more important than in French or English.

```{r read data}
prov_letters <- read_feather(here("Data/providence_letters.feather"))
prov_phones <- read_feather(here("Data/providence_phonemes.feather"))
adult_phonemes <- read_feather(here("Data/adult_phonemes.feather"))
french_letters <- read_feather(here("Data/palasis_letters.feather"))
german_letters <- read_feather(here("Data/wagner_letters.feather"))
```

@@

How much information is in each letter in English words? We use the Providence corpus from CHILDES. In the plots below, each plot represents the information in each letter in words of different lengths by number of letters--we show the length above each plot.

@@

```{r plot prov letter unigrams}
prov_letters %>%
  filter(gram == "Unigram") %>%
  filter(word_length %in% 2:6) %>%
  ggplot(aes(x = letter_order, y = mean_s)) +
    facet_wrap(~word_length) +
    geom_point() +
    geom_line() +
    xlab("Phone Position") +
    ylab("Average Information") + 
    ggtitle("Information in English letters by letter position - unigrams")
```

Now we ask: if you know the last letter, how much information is in the next letter in English? There is relatively less information in each letter knowing the previous letter, but we see the same general information trajectory across letters.

```{r plot prov letter bigrams}
prov_letters %>%
  filter(gram == "Bigram") %>%
  filter(word_length %in% 2:6) %>%
  ggplot(aes(x = letter_order, y = mean_s)) +
    facet_wrap(~word_length) +
    geom_point() +
    geom_line() +
    xlab("Phone Position") +
    ylab("Average Information") + 
    ggtitle("Information in English letters by letter position - bigrams")
```

@@

```{r plot prov letter trigrams}
prov_letters %>%
  filter(gram == "Trigram") %>%
  filter(word_length %in% 2:6) %>%
  ggplot(aes(x = letter_order, y = mean_s)) +
    facet_wrap(~word_length) +
    geom_point() +
    geom_line() +
    xlab("Phone Position") +
    ylab("Average Information") + 
    ggtitle("Information in English letters by letter position - trigrams")
```

@@

```{r plot prov phone unigrams}
prov_phones %>%
  filter(gram == "Unigram") %>%
  filter(word_length %in% 2:6) %>%
  ggplot(aes(x = phone_order, y = mean_s)) +
    facet_wrap(~word_length) +
    geom_point() +
    geom_line() +
    xlab("Phone Position") +
    ylab("Average Information") + 
    ggtitle("Information in English phones by phone position - unigrams")
```

@@

```{r plot prov phone bigrams}
prov_phones %>%
  filter(gram == "Bigram") %>%
  filter(word_length %in% 2:6) %>%
  ggplot(aes(x = phone_order, y = mean_s)) +
    facet_wrap(~word_length) +
    geom_point() +
    geom_line() +
    xlab("Phone Position") +
    ylab("Average Information") + 
    ggtitle("Information in English phones by phone position - bigrams")
```

```{r plot prov phone trigrams}
prov_phones %>%
  filter(gram == "Trigram") %>%
  filter(word_length %in% 2:6) %>%
  ggplot(aes(x = phone_order, y = mean_s)) +
    facet_wrap(~word_length) +
    geom_point() +
    geom_line() +
    xlab("Phone Position") +
    ylab("Average Information") + 
    ggtitle("Information in English phones by phone position - trigrams")
```

@@

```{r show leave-out-entropy}
split_lengths <- adult_phonemes %>%
  group_by(word_length) %>%
  nest() %>%
  filter(word_length > 1) %>%
  filter(word_length %in% 1:6 ) %>%
  mutate(entropy = map(data, hold_out_entropy))

split_lengths %>%
  select(-data) %>%
  unnest() %>%
  ggplot(aes(x = phone_order, y = empirical_stat)) + 
  facet_wrap(~ word_length, scales = "free") +
  geom_pointrange(aes(ymin = ci_lower, ymax = ci_upper)) + 
  theme_classic() + 
  geom_line() + 
  xlab("Phone position") + 
  ylab("Entropy")
```

@@

@@

```{r plot french letter unigrams}
french_letters %>%
  filter(gram == "Unigram") %>%
  filter(word_length %in% 2:6) %>%
  ggplot(aes(x = letter_order, y = mean_s)) +
    facet_wrap(~word_length) +
    geom_point() +
    geom_line() +
    xlab("Phone Position") +
    ylab("Average Information") + 
    ggtitle("Information in French letters by letter position - unigrams")
```

```{r plot french letter bigrams}
french_letters %>%
  filter(gram == "Bigram") %>%
  filter(word_length %in% 2:6) %>%
  ggplot(aes(x = letter_order, y = mean_s)) +
    facet_wrap(~word_length) +
    geom_point() +
    geom_line() +
    xlab("Phone Position") +
    ylab("Average Information") + 
    ggtitle("Information in French letters by letter position - bigrams")
```

```{r plot french letter trigrams}
french_letters %>%
  filter(gram == "Trigram") %>%
  filter(word_length %in% 2:6) %>%
  ggplot(aes(x = letter_order, y = mean_s)) +
    facet_wrap(~word_length) +
    geom_point() +
    geom_line() +
    xlab("Phone Position") +
    ylab("Average Information") + 
    ggtitle("Information in French letters by letter position - trigrams")
```

@@

```{r plot german letter unigrams}
german_letters %>%
  filter(gram == "Unigram") %>%
  filter(word_length %in% 2:6) %>%
  ggplot(aes(x = letter_order, y = mean_s)) +
    facet_wrap(~word_length) +
    geom_point() +
    geom_line() +
    xlab("Phone Position") +
    ylab("Average Information") + 
    ggtitle("Information in German letters by letter position - unigrams")
```

```{r plot german letter bigrams}
german_letters %>%
  filter(gram == "Bigram") %>%
  filter(word_length %in% 2:6) %>%
  ggplot(aes(x = letter_order, y = mean_s)) +
    facet_wrap(~word_length) +
    geom_point() +
    geom_line() +
    xlab("Phone Position") +
    ylab("Average Information") + 
    ggtitle("Information in German letters by letter position - bigrams")
```

```{r plot german letter trigrams}
german_letters %>%
  filter(gram == "Trigram") %>%
  filter(word_length %in% 2:6) %>%
  ggplot(aes(x = letter_order, y = mean_s)) +
    facet_wrap(~word_length) +
    geom_point() +
    geom_line() +
    xlab("Phone Position") +
    ylab("Average Information") + 
    ggtitle("Information in German letters by letter position - trigrams")
```

>>>>

```{r load-libraries, library, message=F, results='hide', warning=FALSE}
library(tidytext)
library(tidyverse)
library(childesr)
library(glue)
library(feather)
library(here)
library(entropy)
library(tidyboot)
library(janitor)
library(arrangements)
library(stringr)

knitr::opts_chunk$set(echo = TRUE, cache = T, warning = F, 
                      fig.width = 8, fig.height = 4)

theme_set(theme_classic(base_size = 16))
```

Where is the information in English words? A salient feature of African American Enlgish Vernacular is the reduction of final consonants and consonant clusters. Why must this happen? One possibility is that the information important for identifying a word is not uniformly distributed across it's characters or sounds but instead concentrated at the beginnings or middles of words. In these cases, reducing final sounds might reduce the articulatory complexity of words without much cost for the listener in processing these words and inferring their intended meaning.

In a series of initial analyses we tackle three ways:
1. Computing the surprisal of characters in English words either from baserates of character production (unigram) or conditioning on the prior charaters in each word (bigram and trigram). This analyses roughly captures the difficulty that a listener might have in predicting each character.
2. Repeating this analysis at the level of phonemes rather than characters, arguably a better proxy for the speaker's and listener's representations
3. Computing hold-out entropy for individual phonemes in English words. This is roughly equivalent to asking "if you knew every other sound in a word, how much trouble would you have predicting what sounds appears in position X"
4. Comparing English to French and German which have different expected information profiles (more below).

SOME KIND OF PREVIEW OF RESULTS HERE?

```{r read-data}
prov_letters <- read_feather(here("Data/providence_letters.feather")) %>%
  mutate(rep = "letter", language = "English") %>%
  rename(order = letter_order)

prov_phones <- read_feather(here("Data/providence_phonemes.feather")) %>%
  mutate(rep = "phone", language = "English") %>%
  rename(order = phone_order)

adult_phonemes <- read_feather(here("Data/adult_phonemes.feather"))

french_letters <- read_feather(here("Data/palasis_letters.feather")) %>%
  mutate(rep = "letter", language = "French") %>%
  rename(order = letter_order)

german_letters <- read_feather(here("Data/wagner_letters.feather")) %>%
  mutate(rep = "letter", language = "German") %>%
  rename(order = letter_order)

all_data <- bind_rows(prov_letters, prov_phones, french_letters, german_letters)
```

@@

We use for our English corpus the Providence corpus from CHILDES. In the plots below, each plot represents the information in each letter in words of different lengths by number of letters--we show the length above each plot.

@@

```{r plot-helper}
plot_curves <- function(select_language = "English", select_rep = "letter", 
                        select_gram = "Unigram") {
  all_data %>%
    filter(language == select_language, rep == select_rep,
           gram == select_gram) %>%
    filter(word_length %in% 2:6) %>%
    ggplot(aes(x = order, y = mean_s)) +
    facet_grid(. ~ word_length, 
               labeller = as_labeller(function(x) paste0("Length ", x))) +
    geom_point() +
    geom_line() +
    xlab(paste0(str_to_sentence(select_rep), " Position")) +
    ylab("Surprisal") + 
    ggtitle(paste0("Information in ", select_language, " ", 
                   select_rep, "s by ",select_rep, " position - ", select_gram))
}
```

```{r prov-letter-unigrams}
plot_curves()
```

Now we ask: if you know the previous letter, how much information is in the next letter in English? There is relatively less information in each letter knowing the previous letter, but we see the same general information trajectory across letters.

```{r prov-letter-bigrams}
plot_curves(select_gram = "Bigram")
```

@@

```{r prov-letter-trigrams}
plot_curves(select_gram = "Trigram")
```

@@

```{r prov-phone-unigrams}
plot_curves(select_rep = "phone")
```

@@

```{r prov-phone-bigrams}
plot_curves(select_rep = "phone", select_gram = "Bigram")
```

```{r prov-phone-trigrams}
plot_curves(select_rep = "phone", select_gram = "Trigram")
```

@@

```{r show-leave-out-entropy}
split_lengths <- adult_phonemes %>%
  group_by(word_length) %>%
  nest() %>%
  filter(word_length > 1) %>%
  filter(word_length %in% 1:6 ) %>%
  mutate(entropy = map(data, hold_out_entropy))

split_lengths %>%
  select(-data) %>%
  unnest() %>%
  ggplot(aes(x = phone_order, y = empirical_stat)) + 
  facet_wrap(~ word_length, 
               labeller = as_labeller(function(x) paste0("Length ", x)),
               scales = "free") +
  geom_pointrange(aes(ymin = ci_lower, ymax = ci_upper)) + 
  theme_classic() + 
  geom_line() + 
  xlab("Phone position") + 
  ylab("Entropy")
```

@@

We are interested in French because when many French words are pronounced, the final letters are not pronounced in any way (e.g. "ils sauraient"="they would know", the final "ent" in "sauraient" is not pronounced at all). We are interested in German because German has a much richer case system than English or French, and case markings in German are suffixes (at the ends of words). So in German, the ends of words may be much more important than in French or English.

@@

```{r plot french letter unigrams}
plot_curves(select_language = "French")
```

```{r plot french letter bigrams}
plot_curves(select_language = "French", select_gram = "Bigram")
```

```{r french-letter-trigrams}
plot_curves(select_language = "French", select_gram = "Trigram")
```

@@

```{r plot german letter unigrams}
plot_curves(select_language = "German")
```

```{r plot german letter bigrams}
plot_curves(select_language = "German", select_gram = "Bigram")
```

```{r plot german letter trigrams}
plot_curves(select_language = "German", select_gram = "Trigram")
```
